Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-07-07 manuscript/manuscript.tex
Finish conclusion

This section identifies advantages of performant models for assigning CODs. Section \ref{discuss-adv-cod} details the strategic use of models for particular CODs. Section \ref{discuss-adv-efficient} details the resource efficiency of GPT and InSilicoVA models for assisting in physician COD assignment. Section \ref{discuss-adv-nlf} notes the strength of using natural language text in GPT models compared to structured data (e.g. questionnaires) for physician COD assignment.

@@

    Injuries & 135 (5\%) & GPT-4 & 0.95\\
    Diarrhoeal diseases & 175 (7\%) & GPT-4 & 0.91\\
    Epilepsy, leukaemia, and\\other noncommunicable diseases & 114 (4\%) & GPT-3.5 & 0.89\\
    Penumonia & 81 (3\%) & InSilicoVA & 0.8\\

@@

\subsubsection{Resource Efficient}\label{discuss-adv-efficient}

The models in this study can assist physicians in assigning CODs in a variety of ways due to low costs and speed of COD assignment. Similar to differential diagnoses, GPT and InSilicoVA models can offer more alternative COD assignments for physicians to consider, \cite{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement in physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA had no costs on 6939 physician agreed records \cite{openaiPricing2024}. These costs are likely lower than physicians (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), while over 10,000 records can be coded in under a day. When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, care needs to be taken to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-perf-geq80}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention For example, physicians can simply check cases, such as road traffic injury where models performed well at 0.84 PCCC, while spending more time on cases that related to acute respiratory infections, where the models performed poorly at 0.62 PCCC (Figure \ref{fig-perf-cod-adult}).

@@

Another research opportunity is in the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-efficient}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting model suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions (Figure \ref{fig-discuss-phyassist}). Our future work will assess the effects of these model suggestions on physician assignment outcomes (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our HEAL-SL study for after survey round 2 \cite{njalauniversityHealthySierraLeone2023}.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.98\textwidth]{fig-discuss-phyassist.pdf}
    \caption{Addition of model suggestions into physician assignment process. GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option to revise or proceed with their initial assignment. The additional consideration potentially reduces disagreement between physicians and ill-defined deaths.}\label{fig-discuss-phyassist}
\end{figure}

@@

TO DO.

@@

The datasets supporting the conclusions of this article are included within the article (and its additional files) and on \url{https://openmortality.org}.

@@

The R and Python code is available on \url{https://github.com/cghr-toronto/rapid-paper1}.

>>>>

This section identifies advantages of performant models for assigning CODs. Section \ref{discuss-adv-cod} details the strategic use of models for particular CODs. Section \ref{discuss-adv-hscav} details the resource efficiency of GPT and InSilicoVA models for assisting in physician COD assignment. Section \ref{discuss-adv-nlf} notes the strength of using natural language text in GPT models compared to structured data (e.g. questionnaires) for physician COD assignment.

@@

    Injuries & 135 (5\%) & GPT-4 & 0.95\\
    Diarrhoeal diseases & 175 (7\%) & GPT-4 & 0.91\\
    Epilepsy, leukaemia, and\\other noncommunicable diseases & 114 (4\%) & GPT-3.5 & 0.89\\
    Pneumonia & 81 (3\%) & InSilicoVA & 0.8\\

@@

\subsubsection{Highly Scalable and Available}\label{discuss-adv-hscav}

The models in this study can assist physicians in assigning CODs in a variety of ways due to low costs and speed of COD assignment. Similar to differential diagnoses, GPT and InSilicoVA models can offer more alternative COD assignments for physicians to consider, \cite{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement in physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA had no costs on 6939 physician agreed records \cite{openaiPricing2024}. These costs are likely lower than physicians (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), while over 10,000 records can be coded in under a day. When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, care needs to be taken to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-perf-geq80}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention For example, physicians can validate cases, such as road traffic injury where models performed well at 0.84 PCCC, while spending more time on cases that related to acute respiratory infections, where the models performed poorly at 0.62 PCCC (Figure \ref{fig-perf-cod-adult}). Lastly, as models can assign CODs on-demand, there is potential for models to provide CODs during the data collection process, and for GPT models to guide the surveyor to ask additional questions for improving narrative quality.

@@

Another research opportunity is in the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-hscav}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting model suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions (Figure \ref{fig-discuss-phyassist}). Our future work will be a first step in computer assisted verbal autopsy, assessing the effects of these model suggestions on physician assignment outcomes (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our HEAL-SL study for after survey round 2 \cite{njalauniversityHealthySierraLeone2023}.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.98\textwidth]{fig-discuss-phyassist.pdf}
    \caption{Addition of model suggestions into the physician assignment process. At step 2, GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option in step 2b to revise or proceed with their initial assignment. The additional consideration potentially reduces disagreement between physicians and ill-defined deaths.}\label{fig-discuss-phyassist}
\end{figure}

@@

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models had CSMF accuracies of 0.7-0.79. At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.45 PCCC). When evaluating performance by COD, 7 of 19, 4 of 10, and 1 of 7 adult (12-69 years), child (28 days to 11 years),
and neonatal (under 28 days) CODs respectively had models with $\geq$0.8 PCCC (see Table \ref{tab-perf-geq80}). Performance decreased from 0.72 to 0.59 PCCC as adults aged, and increased from 0.5 to 0.7 PCCC as children and neonates developed. Thus, GPT and InSilicoVA models were comparable to physicians for particular CODs (e.g. maternal conditions, injuries, diarrhoeal diseases, pneumonia), but not across age ranges. Advantages of GPT (and InSilicoVA for some CODs) models include being highly scalable and available, which allows the suggestion of multiple alternative COD assignments, and reduction of time spent on non-complex cases to assist physicians in COD assignment. In addition, GPT models provide flexible natural language input and output, capturing latent patterns (e.g. health-seeking and social issues) that potentially lead to their overall high performance compared to InterVA-5 and InSilicoVA. However, GPT models do not always assign the same COD for the same record on multiple runs, are trained on data from a specific time period, and require large computing infrastructure, leading to disadvantages in reliability of COD assignments, timeliness of up-to-date information, and data privacy issues. Limitations of this study were small sample sizes in relation to rarer CODs and other national level studies, need for more comprehensive exploration of disagreed and misclassified records, and lack of experimentation of model parameters, output consistency, and multiple COD assignments. Research opportunities include refining GPT models using prompt engineering and custom models, and future work towards computer assisted verbal autopsy, where GPT and other models are used to assist physician COD assignment by offering multiple alternative assignments, improving physician agreement on COD assignment and reducing ill-defined deaths.

@@

The datasets supporting the conclusions of this article are included within the article (and its additional files), at \url{https://openmortality.org} (available upon request) and at \url{https://github.com/cghr-toronto/healsl-gpt-paper}.

\noindent\\Verbal Autopsy (VA) data by age group and survey rounds 1 and 2:

\begin{itemize}
    \item Adult VA (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_adult_v1}
    \item Adult VA (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_adult_v1}
    \item Child VA (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_child_v1}
    \item Child VA (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_child_v1}
    \item Neonatal VA (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_neo_v1}
    \item Neonatal VA (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_neo_v1}
\end{itemize}

\noindent\\Narrative data by age group and survey rounds 1 and 2:

\begin{itemize}
    \item Adult Narratives (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_adult_narrative_v1}
    \item Adult Narratives (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_adult_narrative_v1}
    \item Child Narratives (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_child_narrative_v1}
    \item Child Narratives (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_child_narrative_v1}
    \item Neonatal Narratives (Round 1):\\\url{https://openmortality.org/data/healsl_rd1_neo_narrative_v1}
    \item Neonatal Narratives (Round 2):\\\url{https://openmortality.org/data/healsl_rd2_neo_narrative_v1}
\end{itemize}

\noindent\\Cause of death code mappings to convert between ICD-10, WVA-2016, and CGHR-10 codes:

\begin{itemize}
    \item ICD-10 to CGHR-10:\\\url{https://openmortality.org/data/cghr10_v1}
    \item WVA-2016 to ICD-10:\\\url{https://openmortality.org/data/icd10_wva2016_v1}
\end{itemize}

\noindent\\Result files:

\begin{itemize}
    \item Model and physician COD assignments:\\\url{https://github.com/cghr-toronto/healsl-gpt-paper/blob/main/data/healsl_rd1to2_cod_v1.csv}
    \item Metrics for model COD assignments:\\\url{https://github.com/cghr-toronto/healsl-gpt-paper/blob/main/data/healsl_rd1to2_metrics_v1.csv}
    \item GPT-3.5 repeated runs results in Appendix \ref{app-reprod}:\\\url{https://github.com/cghr-toronto/healsl-gpt-paper/blob/main/data/healsl_rd1to2_rapid_gpt3_sample100_v2b.csv}
\end{itemize}

@@

All code for this paper is available at \url{https://github.com/cghr-toronto/healsl-gpt-paper}.
